
# Quick start {#intro}

*Natália Martínková and Stuart J. E. Baird*

The package `diemr` incorporates the diagnostic index expectation maximisation algorithm used to 
estimate which genomic alleles belong to which side of a barrier to geneflow. To start 
using `diemr`, load the package or install it from CRAN if it is not yet available:

```{r, setup, eval = FALSE}
# Attempt to load a package, if the package was not available, install and load
if(!require("diemr", character.only = TRUE)){
    install.packages("diemr", dependencies = TRUE)
    library("diemr", character.only = TRUE)
}
```
Next, assemble paths to all files containing the data to be used by `diemr`. 
Here, we will use a tiny example dataset that is included in the package for illustrating 
the analysis workflow. A good practice is to check that all files contain data in the 
correct format for all individuals and markers. Additionally, the analysis will need a 
list with **ploidies for all genomic compartments and individuals**, and a vector with 
**indices of samples** that will be included in the analysis.


```{r, eval = FALSE}
filepaths <- system.file("extdata", "data7x3.txt", package = "diemr")
# Analysing six individuals
samples <- 1:6
# Assuming diploid markers of all individuals
nIndividuals <- nchar(readLines(filepaths[1])[1]) - 1)
ploidies = rep(list(rep(2, nIndividuals), length(filepaths))
CheckDiemFormat(files = filepaths, 
                ChosenInds = samples,
                ploidy = ploidies)
# File check passed: TRUE
# Ploidy check passed: TRUE
```

If the `CheckDiemFormat()` function fails, work through the error messages and fix 
the stored input files accordingly. The algorithm repeatedly accesses data from the 
harddisk, so seeing the passed file and variable check prior to the analysis is critical.

To estimate the marker polarities, their diagnostic indices and support, run the function 
`diem()` with default settings. Here, we have only one file with data, so paralelisation 
is unnecessary, and we set `nCores = 1`. **The Windows operating system only allows for 
nCores = 1**. Other operating systems can process multiple genomic compartments (e.g. 
chromosomes) in parallel, the analysis of different genomic compartment files running on 
multiple processors.

```{r, eval = FALSE}
res <- diem(files = filepaths, 
            ploidy = ploidies, 
            markerPolarity = list(c(FALSE, FALSE, TRUE)),
            ChosenInds = samples, 
            nCores = 1)
```

The result is a list, where the element `res$DI` contains a table with marker 
polarities, their diagnostic indices and support. 

```{r, eval = FALSE}
res$DI
#   newPolarity         DI   Support Marker
# 1       FALSE  -4.872256 15.930181      1
# 2        TRUE  -3.520647 18.633399      2
# 3        TRUE -13.274822  6.130625      3
```

The column `newPolarity` means that marker 1 should be imported for subsequent 
analyses as is, and markers 2 and 3 should be imported with 0 replaced with 2 and 2 
replaced with 0 (hereafter 'flipped' \(0 \leftrightarrow 2\)). The marker 3 has the lowest diagnostic 
index and low support, indicating 
that the genotypes scored at this marker are poorly related to the barrier to geneflow 
arbitrated by the data.

With the marker polarities optimised to detect a barrier to geneflow, a plot of the 
polarised genome will show how genomic regions cross the barrier. First, the genotypes
need to be imported with optimal marker polarities. Second, individual hybrid indices 
need to be calculated from the polarised genotypes. And last, the data will be plotted
as a raster image.

```{r, eval = FALSE}
genotypes <- importPolarized(file = filepaths, 
                             changePolarity = res$markerPolarity, 
                             ChosenInds = samples)
                             
HI <- hybridIndex(res$I4)
           
plotPolarized(genotypes = genotypes,
              HI = HI[samples])
```
```{r, echo = FALSE, fig.width = 6, fig.height = 4.5, fig.align = 'center'}
genotypes <- diemr::importPolarized(file = system.file("extdata", "data7x3.txt",
                         package = "diemr"), 
                             changePolarity = c(FALSE, TRUE, TRUE), 
                             ChosenInds = 1:6)
h <- c(.5,0,.33,.5,.83,1)
diemr::plotPolarized(genotypes = genotypes,
              HI = h, useRaster = FALSE)
```

::: {.box .caution}
**CAUTION:** This is just a *quick start* to get you started! For real datasets you will use the diagnostic index (DI) to filter the full set of sites you have analysed with `diem` in order to plot only those markers relevant to any barrier detected in the analysis. See **Chapter \@ref(DIfiltering)** for guidelines.
:::


# Input data {#diemFormat}

*Natália Martínková and Stuart J. E. Baird*

The `diemr` package uses a consise genome representation. Let's have a small dataset 
of three markers genotyped for seven individuals.

```{r, eval = FALSE}
S0011222
S1210001
S02221U0
```
The genotypes encoded as `0` represent homozygotes for an allele attributed to barrier side A, `1` are 
heterozygous genotypes, `2` are homozygotes for another allele, attributed to barrier side B, and `U` 
(symbol "_" is also allowed) represents an unknown state or a third (fourth) allele. The power of 
`diem` lies in the assurance that the user does not need to determine the true assignment to barrier sides A and B before the analysis and the specific genotypes encoded as `0` and `2` respectively 
can be arbitrary.

The leading `S` on each line of the input file is ensures that the marker genotypes are 
read in as a string on all operating systems. The `S` is dropped during import of the genotypes, and 
the dataset is imported as a character matrix of all sites.




## Multiple compartments with different ploidies {#ploidy}

Some genomic compartments differ between individuals in their ploidy. For example, markers located on 
chromosome X in mammals will be diploid in females, but haploid in males. Ploidy differences between 
individuals influence calculation of the hybrid index, which in turn has an effect on the _diem_ analysis.

To set up the _diem_ analysis with multiple compartments, the markers with different individual ploidies
must be stored in separate files. The file analysed in the [**Quick start**](#intro) chapter could contain
markers from autosomes and an additional file will contain markers from an X
chromosome, with individuals 2 and 6 being males. The respective ploidies for the second genomic 
compartment
will be `c(2, 1, 2, 2, 2, 1, 2)`. 

Arguments `files` and `ploidy` will need to reflect the information, taking care that the order of filenames
corresponds to the order of elements in the list of ploidies. `diem` cannot check that the order of the 
elements
is correct, only that the information is in the correct format.

```{r, eval = FALSE}
filepaths2 <- c(system.file("extdata", "data7x3.txt", package = "diemr"),
                system.file("extdata", "data7x10.txt", package = "diemr"))
               
ploidies2 <- list(rep(2, 7),
                  c(2, 1, 2, 2, 2, 1, 2))

CheckDiemFormat(files = filepaths2, 
                ChosenInds = samples,
                ploidy = ploidies2)
# File check passed: TRUE
# Ploidy check passed: TRUE

# Set random seed for reproducibility of null polarities (optional)
set.seed(39583782)

# Run diem 
res2 <- diem(files = filepaths2, 
             ploidy = ploidies2, 
             markerPolarity = FALSE,
             ChosenInds = samples, 
             nCores = 1)
```

Plotting polarised genomes from multiple compartments requires separate import of the compartment data.
The polarities in the `res2$markerPolarity` element are combined across all compartments, and extracting
them requires knowledge of the number of markers in each compartment. Alternatively, the marker polarities
from each compartment can be extracted from the list in the `res2$PolarityChanges` element.

```{r, eval = FALSE}
# Import each compartment into a list
genotypes2 <- importPolarized(file = filepaths2, 
                       changePolarity = res2$markerPolarity, 
                       ChosenInds = samples) 
                       
# Load individual hybrid indices from a stored file
h2 <- hybridIndex("HIwithOptimalPolarities.txt")

# Plot the polarised genotypes
plotPolarized(genotypes = genotypes2,
              HI = h2[samples])
```


```{r, echo = FALSE, warning = FALSE,fig.width = 6, fig.height = 4.5, fig.align = 'center'}
genotypes2 <- Reduce(cbind, list(diemr::importPolarized(file = system.file("extdata", "data7x3.txt", package = "diemr"), 
                       changePolarity = c(FALSE, TRUE, TRUE), 
                       ChosenInds = 1:6),
                  diemr::importPolarized(file = system.file("extdata", "data7x10.txt", package = "diemr"), 
                       changePolarity = c(T,T,T,T,F,T,T,F,T,T), 
                       ChosenInds = 1:6)))
h2 <- c(.53, .217, .571, .39, .35, .85)
diemr::plotPolarized(genotypes = genotypes2,
              HI = h2, useRaster = FALSE)
```

See Chapter \@ref(ploidy2) for extended explanation of ploidy-aware genome polarisation.

